---
title: "Probation Projection-refactored"
date: "January, 2019"
output: 
  html_document:
    code_folding: hide
    
---


```{r setup, include=FALSE}
set.seed(911)
```


## SPAC Illinois Probation Population Projection Model Overview
[... omitted ...]


### Data Sources 
[... omitted ...]


### Required Libraries in R
[... omitted ...]

```{r dependencies, message=FALSE, warning=FALSE, include=FALSE}
install_and_load_pkgs <- function(pkgs) {
  new_pkgs <- pkgs[!pkgs %in% installed.packages()]
  
  if (length(new_pkgs) > 0) install.packages(new_pkgs)
  
  for (pkg in pkgs) {
    library(pkg, character.only = TRUE)
  }
}

pkgs <- c(
  "dplyr",
  "tidyr",
  "lubridate",
  "ggplot2",
  "readr",
  "knitr",
  "rmarkdown",
  "scales",
  "simstudy"
)

install_and_load_pkgs(pkgs)
```


### Retrieve actual AOIC data and simulate CHRI sentence data
[... omitted ...]

```{r simulated_data, echo=TRUE, message=FALSE, warning=FALSE}
## import AOIC data (change to appropriate path)
url <- "https://raw.githubusercontent.com/bobaekang/icjia-r-user-group/master/meeting/20190110_SPAC_projection_example/AOIC_PopulationData.csv"
aoic <- read_csv(url)
names(aoic) <- c("data_year", "circuit", "indic_num", "indic_desc", "caseload")


# define relative frequencies of age and judicial circuit
age_probs <- paste0(
  "0.051;0.052;0.049;0.048;0.046;0.043;0.042;0.039;0.037;0.035;",
  "0.034;0.032;0.03;0.028;0.028;0.026;0.024;0.023;0.022;0.021;",
  "0.019;0.018;0.018;0.017;0.017;0.016;0.016;0.015;0.015;0.014;",
  "0.014;0.013;0.012;0.011;0.01;0.01;0.008;0.007;0.007;0.006;",
  "0.005;0.004;0.003;0.003;0.002;0.002;0.002;0.001;0.001;0.001;",
  "0.001;0;0;0;0;0;0;0;0;0;",
  "0;0;0;0;0;0;0;0;0;0;",
  "0;0;0;0;0;0;0;0;0;0;",
  "0;0;0;0;0"
)

circuit_probs <- paste0(
  "0.026;0.021;0.036;0.035;0.029;0.055;0.029;0.021;0.01;0.037;",
  "0.035;0.024;0.01;0.022;0.022;0.032;0.049;0.079;0.051;0.026;",
  "0.008;0.011;0.022;0.31"
)


# create data definitions for simstudy
dd <- defData(
    varname = "arrest_age",
    dist = "categorical",
    formula = age_probs
  ) %>%
  defData(
    varname = "sent_date",
    dist = "uniformInt",
    formula = "3;2919"
  ) %>%
  defData(
    varname = "circuit",
    dist = "categorical",
    formula = circuit_probs
  ) %>%
  defData(
    varname = "sex",
    dist = "categorical",
    formula = "0.22;0.78"
  ) %>%
  defData(
    varname = "race",
    dist = "categorical",
    formula = "0.357;0.643"
  ) %>%
  defData(
    varname = "sent_length",
    dist = "categorical",
    formula = "0.005;0.174;.105;.524;.139;.023;.029"
  ) %>%
  defData(
    varname = "indic_num",
    dist = "categorical",
    formula = "0.596;0.254;0.150"
  ) %>%
  defData(
    varname = "class4_agg_uuw",
    dist = "categorical",
    formula = "0.95;0.05"
  )


# generate data using data definitions
dx <- genData(222000, dd)
```


### Initial Estimate of the Probation Population Generated from CHRI
[... omitted ...]

```{r adjustment, echo=TRUE, message=FALSE, warning=FALSE}
# define functions
recode_vars <- function(df) {
  df %>%
    mutate(
        sent_date = as.Date("2010-01-01") + sent_date,
        sent_length = if_else(sent_length < 7, sent_length * 180, 1440),
        circuit = if_else(circuit == 24, 88L, circuit),
        indic_num = indic_num + 29,
        class4_agg_uuw = if_else(indic_num > 30, 0, class4_agg_uuw - 1)
      )
}

add_initial_exit_date <- function(df) {
  df %>%
    mutate(
      sent_length_multiplier = 1,
      init_exit_date = sent_date %m+% days(round(sent_length * sent_length_multiplier))
    )
}

add_prev_checks <- function(df, current_year) {
  for (y in 2010:current_year) {
    stock_date_check = as.Date(paste0(y, '/12/31'))
    yr <- paste0('yr_', y)
    df <-
      df %>%
      mutate(
      !!yr := as.numeric(sent_date < stock_date_check & init_exit_date >= stock_date_check)
    )
  }
  df
}

get_latest_year_pop <- function(df, year) {
  df[["final_year"]] <- df[[paste0("yr_", year)]]
  
  df %>%
    group_by(
      data_year = year,
      circuit,
      indic_num
    ) %>%
    summarise(chri_stock_pop = sum(final_year))
}

get_adjust_weight <- function(aoic, year, latest_year_pop) {
  aoic %>%
    filter(data_year == year) %>%
    left_join(latest_year_pop) %>%
    mutate(adjust_weight = caseload / chri_stock_pop)
}

add_adjust_weight <- function(df, aoic, year) {
  latest_year_pop <- get_latest_year_pop(df, year)
  adjust_weight <- get_adjust_weight(aoic, year, latest_year_pop)
	
  left_join(df, adjust_weight)
}


# genearte "ps", a data frame holding the data in the real-data version
latest_year <- 2017

ps <-
  dx %>%
    recode_vars() %>%
    add_initial_exit_date() %>%
    add_prev_checks(latest_year) %>%
    add_adjust_weight(aoic, latest_year)
```

[... omitted ...]

```{r comparison, message=FALSE}
# define functions
get_c_post <- function(ps) {
  ps %>%
    mutate_at(
      vars(matches("yr")),
      funs(. * adjust_weight)
    ) %>%
    select(starts_with("yr")) %>%
    summarise_all(sum) %>%
    gather(
      key = "chri_year",
      value = "post_adjustment"
    ) %>%
    mutate(chri_year = parse_number(chri_year))
}

get_c_pre <- function(ps) {
  ps %>%
    select(starts_with("yr")) %>%
    summarise_all(sum) %>%
    gather(
      key = "chri_year",
      value = "pre_adjustment"
    ) %>%
    mutate(chri_year = parse_number(chri_year))
}

get_aoic_annual_caseload <- function(aoic) {
  aoic %>%
    group_by(data_year) %>%
    summarise(caseload = sum(caseload))
}

get_comparison <- function(ps, aoic_caseload) {
  c_pre <- get_c_pre(ps)
  c_post <- get_c_post(ps)
  
  c_pre %>%
    inner_join(c_post, by = "chri_year") %>%
    left_join(aoic_caseload, by = c("chri_year" = "data_year")) %>%
    mutate(chri_year = as.character(chri_year)) %>%
    select(
      "CY" = chri_year,
      "CHRI Pre-Adjustment" = pre_adjustment,
      "CHRI Post-Adjustment" = post_adjustment,
      "AOIC Caseloads" = caseload
    )
}


# apply functions
aoic_caseload <- get_aoic_annual_caseload(aoic)
comparison <- get_comparison(ps, aoic_caseload)


# show comparison table
comparison %>%
  kable(
    align = "c",
    format.args = list(big.mark = ","),
    caption = paste0(
      "CHRI estimates are simulated from sentences;",
      " AOIC data are from Statistical Reports"
    )
  )
```


### Future Projections
[... omitted ...]

```{r projection, message=FALSE, warning=FALSE}
# define functions
get_latest_stock_pop <- function(ps) {
  ps %>%
    filter(yr_2017 == 1) %>%
    mutate(weight = adjust_weight)
}

get_future_admission <- function(ps, year) {
  filter(ps, year(sent_date) == year)
}

update_future_admission <- function(future_admission, year, trend_pace) {
    future_admission %>%
      mutate(
        sent_date = as.Date(paste(year, month(sent_date), day(sent_date), sep = "/")),
        init_exit_date = sent_date %m+% days(sent_length),
        weight = adjust_weight * trend_pace
      )
  }

update_stock_pop <- function(stock_pop, future_admission, year, sim_num) {
  stock_pop %>%
    bind_rows(future_admission) %>%
    mutate(stock_date_check = as.Date(paste0(year, '/12/31'))) %>%
    filter(init_exit_date >= stock_date_check) %>%
    mutate(mc_sim_num = sim_num)
}

get_single_projection_result <- function(stock_pop) {
  stock_pop %>%
    mutate(data_year = year(stock_date_check)) %>%
    group_by(
      mc_sim_num,
      data_year,
      indic_desc,
      circuit
    ) %>%
    summarise(probation_population = sum(weight))
}

fill_in_projection_result <- function(
  stock_pop,
  future_admission,
  start_year,
  final_year,
  mc_sim_num,
  example
) {
  projection_result <- tibble(
    mc_sim_num = numeric(),
    indic_desc = character(),
    circuit = numeric(),
    probation_population = numeric()
  )
	
  for (year in start_year:final_year) {
    trend_pace <- 1
    if (example == "1PctDecline") {
      trend_pace <- 1 - (year - start_year) / 100
    } else if (example == "montecarlo") {
      trend_pace <- 1 + runif(1,-0.05, 0.05)
    }
    
    future_admission <- update_future_admission(future_admission, year, trend_pace)
    if (example == "Class4AggUUW") {
      future_admission$weight[future_admission$class4_agg_uuw == 1] <- 0
    }
    
    stock_pop <- update_stock_pop(stock_pop, future_admission, year, mc_sim_num)
    
    Remainder <- floor(sum(stock_pop$weight))
    print(paste0(year, " end of year population: ", Remainder))
    
    results <- get_single_projection_result(stock_pop)
    projection_result <- bind_rows(projection_result, results)
  }
  projection_result
}

get_probation_projection <- function(
  ps,
  pull_year,
  start_year,
  final_year,
  mc_sim_num,
  example
) {
  stock_pop <- get_latest_stock_pop(ps)
  future_admission <- get_future_admission(ps, pull_year)
  
  fill_in_projection_result(
    stock_pop,
    future_admission,
    start_year,
    final_year,
    mc_sim_num,
    example
  )
}
```


### Monte Carlo Simulation
[... omitted ...]

```{r montecarlo, warning=FALSE}
# define functions
run_simulation <- function(
  ps,
  pull_year,
  start_year,
  final_year,
  mc_runs,
  example
) {
  df <- tibble(
    mc_sim_num = numeric(),
    indic_desc = character(),
    circuit = numeric(),
    probation_population = numeric(),
    data_year = numeric()
  )
  
  for (i in 1:mc_runs) {
    print(paste0("Begin Simulation #", i))
    
    i_results <- get_probation_projection(
      ps,
      pull_year,
      start_year,
      final_year,
      i,
      example
    )
    
    df <- bind_rows(df, i_results)
  }
  df
}

print_complete_message <- function(start_time, end_time, mc_runs) {
  elapsed_time <- end_time - start_time
  message <- paste0(
    "Simulation complete, performed ",
    mc_runs,
    " simulations in ",
    round(elapsed_time, 2),
    " ",
    attr(elapsed_time, "units"),
    "."
  )
  print(message)
}

get_projection_result <- function(
  ps,
  pull_year,
  start_year,
  final_year,
  mc_runs = 1,
  example = "default"
) {
  start_time <- Sys.time()
  projection_result <-
    run_simulation(ps, pull_year, start_year, final_year, mc_runs, example)
  end_time <- Sys.time()
  print_complete_message (start_time, end_time, mc_runs)
  
  projection_result
}


# set parameter values
pull_year <- latest_year
start_year <- 2018
final_year <- 2035


# run simulation
projection_result <- get_projection_result(
  ps,
  pull_year,
  start_year,
  final_year
)
```

```{r plot, message=FALSE, warning=FALSE}
# define functions
get_historical_projection <- function(projection_result, aoic) {
  projection_result %>%
    mutate(caseload = probation_population) %>%
    bind_rows(aoic) %>%
    group_by(data_year, indic_desc) %>%
    summarise(caseload = sum(caseload))
}

get_historical_projection_caseload <- function(projection_result, aoic) {
  projection_result %>%
    mutate(caseload = probation_population) %>%
    bind_rows(aoic) %>%
    group_by(
      data_year,
      indic_desc,
      circuit
    ) %>%
    arrange(circuit) %>%
    summarise(caseload = sum(caseload)) %>%
    mutate(circuit = if_else(circuit == 88, "Cook", as.character(circuit)))
}


# get historical projections
hp <- get_historical_projection(projection_result, aoic)
hpc <- get_historical_projection_caseload(projection_result, aoic)


# build plots
hp_plot <-
  hp %>%
  ggplot(aes(x = data_year, y = caseload)) +
  geom_line(aes(col = gsub(' [A-z ]*', '' , indic_desc))) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Historical and Projected Probation Populations",
    y = "Caseloads",
    x = "Year",
    color = "Group"
  ) +
  theme_minimal()

hpc_plot <-
  hpc %>%
  ggplot() +
  geom_line(
    aes(
      x = data_year,
      y = caseload,
      col = gsub(' [A-z ]*', '' , indic_desc)
    )
  ) +
  scale_y_continuous(
    labels = scales::comma,
    breaks = scales::pretty_breaks(n = 3)
  ) +
  labs(
    title = "Historical and Projected Probation Populations by Circuit",
    y = "Caseloads",
    color = ""
  ) +
  facet_wrap(~circuit, scales = "free") +
  theme_minimal() +
  theme(
    legend.position = "bottom",
    axis.text.x = element_blank()
  )


# show plots
hp_plot
hpc_plot
```


### Adjustments to Probation Admissions
[... omitted ...]

#### Example - 1% YOY decline, non-compounded
[... omitted ...]

```{r one_pct_decline, message=FALSE, warning=FALSE, echo=TRUE}
# run simulation
projection_result <- get_projection_result(
  ps,
  pull_year,
  start_year,
  final_year,
  example = "1PctDecline"
)

hp_1 <- get_historical_projection(projection_result, aoic)


# show plot
hp_plot +
  geom_line(
    data = hp_1,
    aes(
      x = data_year,
      y = caseload,
      col = gsub(' [A-z ]*', '' , indic_desc),
      linetype = gsub(' [A-z ]*', '' , indic_desc)
    ),
    show_guide = FALSE
  ) +
  labs(
    title = "Status Quo and Projected Probation Populations",
    subtitle = "1% annual decline in admissions",
    y = "Caseloads",
    x = "Year",
    color = "Group"
  )
```


#### Example - Class 4 Aggravated UUW Non-Probationable
[... omitted ...]

```{r class4_agg_uuw, message=FALSE, warning=FALSE, echo=TRUE}
# run simulation
projection_result <- get_projection_result(
  ps,
  pull_year,
  start_year,
  final_year,
  example = "Class4AggUUW"
)

hp_2 <- get_historical_projection(projection_result, aoic)


# show plot
hp_plot +
  geom_line(
    data = hp_2,
    aes(
      x = data_year,
      y = caseload,
      col = gsub(' [A-z ]*', '' , indic_desc),
      linetype = gsub(' [A-z ]*', '' , indic_desc)
    ),
    show_guide = FALSE
  ) +
  labs(
    title = "Status Quo and Projected Probation Populations",
    subtitle = "Making Class 4 Agg UUW Non-Probationable",
    y = "Caseloads",
    x = "Year",
    color = "Group"
  )
```


#### Monte Carlo simulation example
[... omitted ...]

```{r montecarlo_example, message=FALSE, warning=FALSE}
# define functions
get_mc_plot_data <- function(projection_result, aoic_caseload) {
  projection_result %>%
    mutate(
      caseload = probation_population,
      Simulation = as.factor(mc_sim_num)
    ) %>%
    group_by(Simulation, data_year) %>%
    summarise(caseload = sum(caseload)) %>%
    bind_rows(aoic_caseload)
}

get_mc_plot_data2 <- function(projection_result) {
  projection_result %>%
    mutate(caseload = probation_population) %>%
    group_by(mc_sim_num, data_year) %>%
    summarise(caseload = sum(caseload)) %>%
    group_by(data_year) %>%
    summarise(
      min_caseload = min(caseload),
      max_caseload = max(caseload),
      avg_caseload = mean(caseload)
    )
}


# run simulation
projection_result <- get_projection_result(
  ps,
  pull_year,
  start_year,
  final_year,
  mc_runs = 20,
  example = "montecarlo"
)


# apply functions
mc_plot_data <- get_mc_plot_data(projection_result, aoic_caseload)
mc_plot_data2 <- get_mc_plot_data2(projection_result)


# display plots
mc_plot_data %>%
  ggplot(aes(x = data_year, y = caseload)) +
  geom_line(aes(color = Simulation)) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Historical and Projected Probation Populations",
    y = "Caseloads",
    x = "Year"
  )

aoic_caseload %>%
  ggplot(aes(x = data_year, y = caseload)) +
  geom_line() +
  geom_ribbon(
    data = mc_plot_data2,
    aes(
      x = data_year,
      y = avg_caseload,
      ymin = min_caseload,
      ymax = max_caseload
    ),
    fill = "#0000FF"
  ) +
  geom_line(
    data = mc_plot_data2,
    aes(y = avg_caseload),
    color = "red",
    size = 1.5
  ) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Historical and Projected Probation Populations",
    y = "Caseloads",
    x = "Year"
  ) +
  theme(legend.position = "None")
```

### Limitations
[... omitted ...]

### Final Code
```{r showblocks, ref.label=c("setup", "dependencies", "simulated_data", "adjustment", "comparison", "projection", "montecarlo"), eval=FALSE}
```


